Generalists are the most urban-tolerant of birds: a phylogenetically controlled analysis of ecological and life-history traits using a novel continuous urbanization index

### 2018-06-11 ms.Rmd
some edits from RK

# ABSTRACT
Identifying which ecological and life history traits influence a species' tolerance to urbanization is critical to understanding species trajectories of biodiversity in an increasingly urbanizing world. There is evidence for a wide array of contrasting patterns for single trait associations with urbanization. In a continental-scale analysis, incorporating 477 species and > 5,000,000 bird observations, we provided a novel and scalable methodology that evaluated the ecological and life history traits which most influence a species' adaptability to persist in urban environments. We identified generalized, phylogenetically controlled patterns: bird species which are generalists (i.e., large niche breadth), with high clutch size, and large residual brain size are among the most urban-tolerant bird species. Conversely, specialized feeding strategies (i.e., insectivores and granivores) were negatively associated with urbanization. Enhancement and persistence of avian biodiversity in urban environments probably relies on protecting, maintaining, and restoring diverse habitats serving a range of life history strategies.

@@

We filtered the bird observations [@callaghan2017assessing; @la2014role] only including observations on complete checklists and which were either travelling, random, stationary, area, or followed BirdLife Australia protocols [@barrett2003new] (see here: http://help.ebird.org/customer/portal/articles/1006209-understanding-observation-types). We then filtered the checklists to those which travelled < 5 km or covered < 500 Ha. Any checklists shared among multiple individuals were subsampled, including only one checklist in order to avoid duplicate results. This provided 637,482 eligible checklists throughout Australia. Seabirds (gannets, petrels, shearwaters, etc.) were omitted from the analyses as we did not expect any correlation with seabirds using urban areas. Further, species with > 100 observations were considered for further analyses. Totally, 580 species, a substantial proportion of the ~730 bird species which are resident in Australia, were considered for analyses. Given the diverse datasets (see below) used in this analysis, there were differing taxonomic authorities followed (see Appendix 2 for elaboration).

## Urbanization index
We used VIIRS night-time lights [@elvidge2017viirs] as a proxy of a bird's association with urban habitat. The VIIRS night-time lights measures the electric lighting present on Earth's surface which is generally associated with human settlement (Fig. S1). There are also a significant number of steps (e.g., filtering out background noise such as fire, degraded data, and other light source contamination) which are used to ensure the validity of the data and its association with human settlement [@elvidge2017viirs]. For the 637,482 checklists which met our criteria above, we used the checklists' associated spatial coordinates and Google Earth Engine [@gorelick2017google] to assign the average radiance value calculated from the VIIRS layer. For each checklist, the average radiance was calculated within a 5 km buffer, in order to match the spatial scale of the checklists included and to minimize any potential bias in eBird sampling protocols. Any radiance values which were negative were forced to 0.00001 [*sensu* @ou2015evaluation]. Each checklist's average radiance value was then passed to every bird on that checklist, which provided every species with a distribution of average radiance values (Fig. 1). The median for each species' given distribution of average radiance values served as the response variable in our analyses. Inherently, there is no definitive range for this index, as it solely depends on a species' distribution of observations.

## Biological traits
Many biological and life history traits have been used in previous studies, and potential traits in our study were chosen on the basis of their existing support and hypotheses in the current literature [e.g., @mcclure1989characterizes; @kark2007living; @evans2011makes]; we sought to encompass a broad array of traits previously tested. Hence, predictor variables were calculated using a published dataset which contains the biological, ecological, and conservation information for all of Australia's birds [@garnett2015biological]. Table 1 provides a brief overview of the traits that we investigated, while Appendix 1 Appendix 1 provides an overview of all candidate traits, some associated references for each trait, and more detailed information on how the traits were calculated. Traits were parsed in different manners (Appendix 1b), and for any missing value in the dataset, we treated that as evidence against it existing (i.e., if a species had NA for a specific habitat, we assumed that the lack of evidence found by the authors indicates that the species is unlikely to be found in that habitat). Preliminary analyses including IUCN status (exotic/native) showed that exotic birds were strongly associated with urbanization, but given the preferential selection by acclimatization societies for introduction of species with proven synanthropy [@blackburn2009avian], exotic species were omitted from the final analysis to avoid bias. Therefore, we excluded IUCN status as a trait because the majority of our candidate species were of Least Concern. Only species which had complete data for the list of candidate traits were considered in the analyses, and 477 species met this criteria.

## Statistical analyses
All models and analyses were conducted in R [@r2017coreteam] and relied heavily on the tidyverse workflow [@wickham2017tidyverse].

Before modelling, the predictor variables were investigated for multi-collinearity. We used the corrplot package [@corrplot2017] to show correlation among continuous variables (Fig. S2), and used generalized variance inflation factors [@fox1992generalized], from the car package [@fox2011companion], which accounts for the degrees of freedom for a particular term to show minimal correlation amongst each other (Supplementary Table 1).

We employed two linear modelling approaches to provide robustness of our results; one with and one without phylogenetic constraints. Phylogenetic models were used because of the importance of considering evolutionary distinctedness in ecological research [@losos2008phylogenetic; @graham2018phylogenetic], but given that phylogenetically designed analyses can have a number of shortcomings [@cadotte2017phylogenies], we also included non-phylogenetic models in our analysis. Our urbanization index was log-transformed in order to meet model assumptions. The predictor variables used in the models are listed in Table 1.

We weighted the linear regression models [@solon2015weighting] using a source of error associated with our urbanization indices. This error was the proportion of the number of total observations of a species divided by the number of unique localities a species was observed from. This gave more weight in the model-fitting procedure to those species that had more observations in more unique locations. This was done based on the assumption that those species which had the most observations would have the most reliable urbanization indices. This approach also addresses biases in sampling of eBird data in which some birds can have a high number of observations, but from a small number of locations such as known roosting sites [e.g., @callaghan2018avian].

### Non-phylogenetically controlled models
Given that we hypothesized any of the potential candidate traits could significantly drive the response in our urbanization index, we first ran one global model including all 22 predictor variables. We used the standardize function from the arm package [@gelman2008scaling; @gelman2016arm] to standardize model estimates to represent relative effect sizes. In order to confirm our linear model results, we employed a model-averaging approach [@grueber2011multimodel] in which we 'dredged' possible subsets of the global model. This was done using the dredge function in the MuMIn package [@barton2016mumin], and we allowed a total of 11 candidate predictors to be fit at any one time in order to avoid over-fitting. This left a total of 784,626 models which were assessed, and the model-averaged parameter estimates were calculated from the top models of those with $\Delta$AIC < 4. As with the global model, we present the standardized estimates which represent relative effect sizes.

### Phylogenetic analysis
As the correlation between traits and urbanization may be influenced by the shared evolutionary history of the species [@felsenstein1985phylogenies], we also used phylogenetically informed models to test our hypotheses. Phylogenetically informed models are descendants of phylogenetic independent contrasts [@felsenstein1985phylogenies], which in simple terms, tests for repeated evolutionary association between the traits of interest. Or in this particular case, whether the traits of interest (e.g., large brain size at a given body size) repeatedly allow for the use of urban environments wherever it evolved across the bird phylogeny. We used the phylogenetic trees provided by Jetz et al. [@jetz2012global]. This phylogeny represents the first completely mapped phylogeny of 9,993 birds [@jetz2012global]. In order to assess whether there was a phylogenetic signal [@revell2008phylogenetic; @munkemuller2012measure] of the urbanization index and the continuous predictor variables used in our study, we used the phylosignal package [@keck2016phylosignal] in the R computing environment [@r2017coreteam]. This was also conducted for each of the continuous response variables in order to assess the relative contribution of phylogeny on the predictor variables.

Phylogenetic models were fit based on the same global model as above, but using the 'phylolm' function from the phylolm package which fits a linear regression model where the likelihood is linear in the number of tips in the tree [@ho2014phylolm]. Similar to above, we model-averaged the global phylogenetic model using the aforementioned methods. To account for aspects of phylogenetic uncertainty, we also fit the same model to a range of equally likely trees from Jetz *et al.* [-@jetz2012global] examining the distribution of putatively important parameters across the distribution of trees.

# RESULTS
A total of 316,306 checklists resulting in 5,944,819 observations of 477 different species were included in analyses. The least often recorded species was *Turnix melanogaster* (Black-breasted Buttonquail; N=104) while the most recorded species was *Cracticus tibicen* (Australian Magpie; N=158,615). The mean number of observations for a species was 10,250. Plots of all 477 species' distributions of average radiance values are presented in Appendix 3. A complete list of the 477 species, in addition to their traits, number of observations, and their urbanization indices can be found in Appendix 4.

## Urbanization index
For the 477 species included in the analysis the mean $\pm$ SD urbanization index was 0.91 $\pm$ 1.76, with a range from 0.008 (*Cracticus mentalis* Black-backed Butcherbird) to 11.85 (*Ninox strenua* Powerful Owl; Fig. 2).

>>>>

# ABSTRACT
Identifying which ecological and life history traits influence a species' tolerance to urbanization is critical to understanding the trajectory of biodiversity in an increasingly urbanizing world. There is evidence for a wide array of contrasting patterns for single trait associations with urbanization. In a continental-scale analysis, incorporating 477 species and > 5,000,000 bird observations, we provided a novel and scalable methodology that evaluated the ecological and life history traits which most influence a species' adaptability to persist in urban environments. We identified generalized, phylogenetically controlled patterns: bird species which are generalists (i.e., large niche breadth), with high clutch size, and large residual brain size are among the most urban-tolerant bird species. Conversely, specialized feeding strategies (i.e., insectivores and granivores) were negatively associated with urbanization. Enhancement and persistence of avian biodiversity in urban environments probably relies on protecting, maintaining, and restoring diverse habitats serving a range of life history strategies.

@@

We filtered the bird observations [@callaghan2017assessing; @la2014role] only including observations on complete checklists and which were either travelling, random, stationary, area, or followed BirdLife Australia protocols [@barrett2003new] (see here: http://help.ebird.org/customer/portal/articles/1006209-understanding-observation-types). We then filtered the checklists to those which travelled < 5 km or covered < 500 Ha. Any checklists shared among multiple individuals were subsampled, including only one checklist in order to avoid duplicate results. This provided 637,482 eligible checklists throughout Australia. Seabirds (gannets, petrels, shearwaters, etc.) were omitted from the analyses as we did not expect any correlation with seabirds using urban areas. Further, species with > 100 observations were considered for further analyses. Totally, 580 species, a substantial proportion of the ~730 bird species which are resident in Australia, were considered for analyses. We used differing taxonomic authorities (see Appendix 2 for elaboration).

## Urbanization index
We used VIIRS night-time lights [@elvidge2017viirs], for urban habitat, measuring the electric lighting present on Earth's surface, generally associated with human settlement (Fig. S1). The night-time lights dataset filters out background noise (e.g., fires, degraded data, and other light source contamination) to ensure validity of the data and its association with human settlement [@elvidge2017viirs]. For our 637,482 checklists and their associated species, we used the spatial coordinates and Google Earth Engine [@gorelick2017google], assigning the average radiance value calculated from the VIIRS layer. For each checklist, the average radiance was calculated within a 5 km buffer, in order to match the spatial scale of the checklists included and to minimize any potential bias in eBird sampling protocols. Any negative radiance values were forced to 0.00001 [*sensu* @ou2015evaluation]. Every bird species subsequently had a distribution of average radiance values (Fig. 1), with each median used as the response variable in our analysis. There was no definitive range for this index, depending on a species' distribution of observations.

## Ecological and life history traits
Ecological and life history traits were chosen on the basis of existing support [e.g., @mcclure1989characterizes; @kark2007living; @evans2011makes; Table 1]. This included a wide array of potential traits for all of Australia's birds [@garnett2015biological]. Traits included body size, residual brain size, clutch size, breeding habitat generalism, feeding habitat generalism, diet generalism, nesting generalism, range size, nomadic/irruptiveness, whether a species was migratory, Table 1 provides a brief overview of the traits that we investigated, while Appendix 1 Appendix 1 provides an overview of all candidate traits, some associated references for each trait, and more detailed information on how the traits were calculated. Traits were parsed in different manners (Appendix 1b), and for any missing value in the dataset, we treated that as evidence against it existing (i.e., if a species had NA for a specific habitat, we assumed that the lack of evidence found by the authors indicates that the species is unlikely to be found in that habitat). Preliminary analyses including IUCN status (exotic/native) showed that exotic birds were strongly associated with urbanization, but given the preferential selection by acclimatization societies for introduction of species with proven synanthropy [@blackburn2009avian], exotic species were omitted from the final analysis to avoid bias. Therefore, we excluded IUCN status as a trait because the majority of our candidate species were of Least Concern. Only species which had complete data for the list of candidate traits were considered in the analyses, and 477 species met this criteria.

## Statistical analyses
Before modelling of predictor variables (Table 1) we investigated multi-collinearity, using the corrplot package [@corrplot2017], among continuous variables (Fig. S2). We used generalized variance inflation factors [@fox1992generalized], from the car package [@fox2011companion], which accounts for the degrees of freedom for a particular term to show minimal correlation amongst each other, and all traits were < 2 suggesting negligible correlation (Supplementary Table 1). We used two independent linear modelling approaches: one without and one with phylogenetic constraints. Both approaches were used because of the associated shortcomings [@cadotte2017phylogenies] and importance [@losos2008phylogenetic; @graham2018phylogenetic] of incorporating evolutionary distinctness in ecological research.

We weighted the linear regression models [@solon2015weighting] using a source of error associated with our urbanization indices, which were log-transformed to meet model assumptions. This error was the proportion of the number of total observations of a species, divided by the number of its unique localities. This weighted the model-fitting to those species with more observations in more unique locations, likely to be the most reliable urbanization indices. This approach also addressed biases in sampling of eBird data where some birds have a high number of observations from a small number of locations such as known roosting sites [e.g., @callaghan2018avian].

All models and analyses were conducted in R [@r2017coreteam] and relied heavily on the tidyverse workflow [@wickham2017tidyverse].

### Non-phylogenetically controlled models
We first ran one global model, including all 22 predictor variables, using the standardize function from the arm package [@gelman2008scaling; @gelman2016arm] to standardize model estimates to represent relative effect sizes. To corroborate our linear model results, we employed a model-averaging approach [@grueber2011multimodel] where we 'dredged' possible subsets of the global model, using the dredge function in the MuMIn package [@barton2016mumin]. We fitted a total of 11 candidate predictors at any one time to avoid over-fitting, totalling 784,626 models assessed. Model-averaged parameter estimates were calculated from the subsequent top models, with $\Delta$AIC < 4, with standardized estimates representing relative effect sizes.

### Phylogenetic analysis
As associations between traits and urbanization may be influenced by shared evolutionary history of the species [@felsenstein1985phylogenies], we also used phylogenetically informed models. These test for repeated evolutionary association between traits of interest [@felsenstein1985phylogenies]. We used published phylogenetic trees for 9,993 species [@jetz2012global] to assess existence of a phylogenetic signal [@revell2008phylogenetic; @munkemuller2012measure] associated with our urbanization index and for each of the other continuous predictor variables (Table 1). This was to assess the relative contribution of phylogeny on the predictor variables, and we used the phylosignal package [@keck2016phylosignal] in R.

Phylogenetic models were fit based on the same global model as with the non-phylogenetic models, using the 'phylolm' function from the phylolm package which fits a linear regression model where the likelihood is linear in the number of tips in the tree [@ho2014phylolm]. We also model-averaged the global phylogenetic model using the aforementioned methods. To account for phylogenetic uncertainty, we also fitted the same model to a range of equally likely trees from Jetz *et al.* [-@jetz2012global], examining the distribution of putatively important traits across the distribution of trees.

# RESULTS
We analyzed 316,306 checklists including 5,944,819 observations of 477 different species. The least recorded species was *Turnix melanogaster* (Black-breasted Buttonquail; N=104) while the most recorded species was *Cracticus tibicen* (Australian Magpie; N=158,615). The mean ($\pm$ SD) number of observations for a species was 10,250 $\pm$ xxxx. Distributions of measured VIIRS average radiance values (Appendix 3), their traits, number of observations, and urbanization indices for the 477 species (Appendix 4) varied considerably. The mean $\pm$ SD urbanization index was 0.91 $\pm$ 1.76, ranging from 0.008 (*Cracticus mentalis* Black-backed Butcherbird) to 11.85 (*Ninox strenua* Powerful Owl; Fig. 2).
